Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2017-05-13 documentation/userguide.tex
Remove source/ and tests/, these are in Castro now

% codes
\newcommand{\castro}{{\sf Castro}}
\newcommand{\boxlib}{{\sf BoxLib}}
\newcommand{\yt}{{\sf yt}}

@@

Welcome to the user's guide for the \texttt{wdmerger} software package. This software is designed to
simulate binary white dwarf systems, and is intended to provide useful information on the 
viability of mergers of white dwarfs as a progenitor for Type Ia supernovae. It is a set of 
source files designed for use with the \texttt{CASTRO} code \citep{castro}.

This software requires a few other software packages to be installed before it can be used.
The science software includes \href{http://boxlib-codes.github.io/}{BoxLib} and 
\href{http://boxlib-codes.github.io/Castro/}{CASTRO}. You should also 
download the \href{https://github.com/BoxLib-Codes/Microphysics}{Microphysics} repository, 
which stores the equation of state and nuclear network we will use. This software is written 
in C++ and Fortran, and compilers for each are needed. The gcc compilers are sufficient.
In addition, it relies on several utilities that are 
commonly distributed on Linux systems: \texttt{bash}, \texttt{GNU bc}, \texttt{GNU make}, 
\texttt{batch}, \texttt{GNU sed}, and \texttt{GNU awk}. Analysis is mainly performed in 
\href{https://www.python.org/}{python}, with the \href{http://matplotlib.org/}{matplotlib} 
library and the \href{http://yt-project.org/}{yt} code being used for visualization. 
If python and yt are not installed on your system, you will still be able to compile and 
run this code, but we use yt for all analysis of \texttt{BoxLib} output data other than 
the ASCII files that are generated at runtime. We have some routines for this purpose 
available in Fortran and Python; some of the Fortran tools are provided in \texttt{BoxLib} 
itself, in the \texttt{Tools/Postprocessing} subdirectory, and it is straightforward enough 
to modify them for basic analysis. We may be able to provide others upon request.

This software assumes that you have set several environment variables:
\begin{itemize}
  \item \texttt{BOXLIB\_HOME}, the location of the \texttt{BoxLib} code's top-level directory
  \item \texttt{CASTRO\_DIR}, the location of the \texttt{CASTRO} code's top-level directory
  \item \texttt{WDMERGER\_HOME}, the location of this software's top-level directory
  \item \texttt{MICROPHYSICS\_DIR}, the location of the \texttt{Microphysics} code's top-level directory
\end{itemize}

@@

The \texttt{source} directory contains all of the Fortran routines needed to build a \texttt{CASTRO} 
executable file. You can type \texttt{make} in this directory to build a vanilla executable 
with the default settings, though usually we will build the software from a different location.

The \texttt{analysis} directory contains the Python analysis scripts used for analyzing output
data (both in \texttt{BoxLib} and plain-text format).

@@

The \texttt{tests} directory is designed to hold small versions of problems that can be run in a
nightly test suite that regularly checks for code errors.

@@

This chapter describes the setup of the white dwarf merger problem. It includes descriptions of 
the physics involved, the code options and runtime parameters, the diagnostic output 
and other I/O information, and the analysis routines. A full description of the physics 
and the associated numerical algorithms is included in the \texttt{CASTRO} code paper \cite{castro}.

@@

\texttt{CASTRO} includes full self-gravity with the Poisson equation, as well as the ability to perform a
simulation in a rotating reference frame with fixed rotation period. By default the $z$ axis will 
be the rotation axis, with the stars initially aligned along the $x$ axis. We select the 
orbital period of the system, and the masses of the \textit{primary} (more massive) and 
\textit{secondary} (less massive) white dwarfs. We then use Kepler's third law to determine the 
appropriate initial orbital distance. The center of mass of the system corresponds with the center 
of the domain, which is the origin; the domain is $1.024 \times 10^{10}\ \text{cm}$ wide in each 
coordinate direction, so that each dimension runs from $-5.12 \times 10^{9}\ \text{cm}$ to 
$+5.12 \times 10^{9}\ \text{cm}$. In a rotating reference frame, we set the stars at rest. 
We can instead run a simulation in the inertial reference frame. To make it equivalent, we
start all zones in the domain with a velocity corresponding to the equivalent rotating 
reference frame velocity. This means that the stars will be synchronously rotating (though we
may relax this assumption in future work).

Though we are mainly interested in 3D simulations, it is sometimes useful to compare to 2D
simulations which are much cheaper and can be run at higher resolutions. The only physically
meaningful analogue to these mergers in lower dimensions occurs in 2D axisymmetric cylindrical
coordinates. CASTRO represents these as $(r,z,\theta)$. In this coordinate system we enforce
that the stars lie along the $z$ axis and that the center of the stars lies at $r = 0$. The
$\theta$ dimension doesn't get directly evolved hydrodynamically but that is the dimension
along which the rotational motion occurs, so in some circumstances the velocity in that
direction will change with time because we still allow the $\theta$-velocity to participate in
the Coriolis force. When rotation is enabled, the rotation is along the $z = 0$ axis in the $r$
coordinate dimension. To try this out you should use \texttt{DIM = 2} in the makefile and
use \texttt{inputs\_2d} in the \texttt{source/} directory.

@@

There are a few \texttt{CASTRO} input parameters that you will typically want to control when running problems. 
Among the most important is whether we are doing rotation (\texttt{castro.do\_rotation}) and if so, 
what the rotational period is (\texttt{castro.rotational\_period}). (Note that the rotational period 
is used to set the orbital properties even if we're not rotating, so you should always set this appropriately.)
You'll also want to change the \texttt{stop\_time}, which is when the simulation will terminate. 
You should also avoid flooding yourself with plotfiles and checkpoints by setting \texttt{amr.plot\_files\_output} and 
\texttt{amr.checkpoint\_files\_output}, respectively; these are the simulation time differences
between successive plotfiles and checkpoints. Note that \texttt{CASTRO} does not adjust the timestep 
to meet this criterion, so it will generate these at the first timestep that passes this interval.

@@

examining \texttt{source/probin}. (Some parameters in there are currently beta quality and 

@@

The \texttt{tagging} namelist is described in the AMR chapter of the \texttt{CASTRO} user's guide. At present, we 
only tag based on absolute values of the density and temperature.

@@

The routines that generate both of these are found in \\\texttt{source/sum\_integrated\_quantities.cpp}. 
This is a replacement of the \texttt{CASTRO} function that does something similar. We use the same 
conventions; for example, if summing over the whole grid and we encounter a region that is covered by 
a fine grid, we zero out the contribution from the coarse grid and only count the fine grid quantities 
in that location.

@@

At a minimum, your working directory needs a \texttt{compile/} subdirectory with a \texttt{makefile}
in it, and some \texttt{bash} script that you run to submit the jobs. If you want to build 
with the default settings, then your makefile needs only to have the line:
\begin{lstlisting}
  include $(WDMERGER_HOME)/source/GNUmakefile
\end{lstlisting}%$
This includes the main makefile from the \texttt{source/} directory in \texttt{wdmerger}. 
The main makefile sets all of our build options (compiler, number of space dimensions, 
equation of state, etc.) and then includes the \texttt{CASTRO} makefile, which then runs as normal.
If you want to adjust the build options, then simply add a line before this include. As an 
example, if you want to build using a gamma-law equation of state, set:
\begin{lstlisting}
  EOS_dir = gamma_law_general/

@@

Decide what directory you want to build the problem in. It is expected 
to be a subdirectory of the \texttt{results/} directory, which will be
automatically created for you:
\begin{lstlisting}
dir=results/test_problem/
\end{lstlisting}
Now, optionally decide how many processors you want to run with, and 
how long you want the job to run:
\begin{lstlisting}
nprocs=128
walltime=4:00:00
\end{lstlisting}
If you do not, the job submission script will default to running the job
using one node for one hour. Finally, run your job:
\begin{lstlisting}
run $dir $nprocs $walltime
\end{lstlisting}%$
That is all you need to submit a job. Return to the terminal, and type \texttt{bash run.sh}
and your job will execute. The helper function \texttt{run} does several things. First, 
it creates the directory named \texttt{\$dir}. Then, it copies into \texttt{\$dir} the \texttt{inputs}
and \texttt{probin} files from the root \texttt{wdmerger} source directory; these are the runtime
files \texttt{CASTRO} needs to determine what your problem setup will be. They contain the default 
parameters for merger runs. It also copies any auxiliary files one might need, like the 
table for the equation of state, as well as the actual \texttt{CASTRO} executable you just built in the 
\texttt{compile/} directory, and a copy of the run utilities that manage the job itself (so that the 
run doesn't change if you make local edits to the job management system). Finally, it builds a script 
that you'll use for actually submitting the job to the batch submission system on the machine you're on, 
and then submits that job. Thus for a typical setup, your run directory will contain at least the following 
files when you submit:
\begin{itemize}
  \item \texttt{Castro3d*.ex}
  \item \texttt{helm\_table.dat}
  \item \texttt{inputs}
  \item \texttt{probin}
  \item \texttt{run\_script}
  \item \texttt{job\_scripts/}
\end{itemize}
Note that if you want to set up this directory as above but \textit{not} actually submit the job, 
set the \texttt{no\_submit} variable to any value in your run script. Similarly, if you want to 
set up the directory to contain the inputs and probin file but none of the machine-dependent 
scripts or the executable, set the \texttt{inputs\_only} variable to any value in your run script.

@@

The \texttt{science/wdmerger\_I/} directory contains a number of example problems that we use
to assess the reliability of \texttt{CASTRO} in performing hydrodynamics simulations 
with self-gravity. This section includes a list of the existing problems 
as well as the instructions for building and running a problem.

@@

This project was started as the Ph.D. dissertation project for Max Katz at Stony Brook University. 
Michael Zingale is Max's primary thesis advisor. Correspondence and requests for information 
about this software should be directed to them; their e-mail addresses are maximilian.katz@stonybrook.edu
and michael.zingale@stonybrook.edu, respectively. Alan Calder and Doug Swesty at Stony Brook are 
also co-investigators on this project. The \texttt{CASTRO} code was developed primarily at the 
Lawrence Berkeley National Laboratory. Staff there that have contributed directly and 
indirectly to this project include Ann Almgren, John Bell, Weiqun Zhang, Andy Nonaka, and Vince Beckner.

>>>>

% codes
\newcommand{\castro}{{\sf Castro}}
\newcommand{\amrex}{{\sf AMReX}}
\newcommand{\yt}{{\sf yt}}

@@

Welcome to the user's guide for the \texttt{wdmerger} software package. This software is designed to
simulate binary white dwarf systems, and is intended to provide useful information on the 
viability of mergers of white dwarfs as a progenitor for Type Ia supernovae. It is a set of 
driver routines and analysis scripts designed for use with the \castro\ code \citep{castro}.

This software requires a few other software packages to be installed before it can be used.
The science software includes \href{https://github.com/AMReX-Codes/amrex}{AMReX} and 
\href{https://github.com/AMReX-Astro/Castro}{CASTRO}. \castro\ will also need
the \href{https://github.com/BoxLib-Codes/Microphysics}{Microphysics} repository, 
which stores the equation of state and nuclear network we will use. Instructions for
building and installing \castro\ can be found in its documentation.
This package relies on several utilities that are
commonly distributed on Linux systems: \texttt{bash}, \texttt{GNU bc}, \texttt{GNU make}, 
\texttt{batch}, \texttt{GNU sed}, and \texttt{GNU awk}. Analysis is mainly performed in 
\href{https://www.python.org/}{python}, with the \href{http://matplotlib.org/}{matplotlib} 
library and the \href{http://yt-project.org/}{yt} code being used for visualization. 
If python and yt are not installed on your system, you will still be able to compile and 
run this code, but we use yt for all analysis of \amrex\ output data other than 
the ASCII files that are generated at runtime. We have some routines for this purpose 
available in Fortran and Python; some of the Fortran tools are provided in \amrex\ 
itself, in the \texttt{Tools/Postprocessing} subdirectory, and it is straightforward enough 
to modify them for basic analysis. We may be able to provide others upon request.

This software assumes that you have set several environment variables:
\begin{itemize}
  \item \texttt{AMREX\_HOME}, the location of the \amrex\ code's top-level directory
  \item \texttt{CASTRO\_DIR}, the location of the \castro\ code's top-level directory
  \item \texttt{WDMERGER\_HOME}, the location of this software's top-level directory
  \item \texttt{MICROPHYSICS\_DIR}, the location of the \texttt{Microphysics} code's top-level directory
\end{itemize}

@@

The \texttt{analysis} directory contains the Python analysis scripts used for analyzing output
data (both in \amrex\ and plain-text format).

@@

@@

This chapter describes the setup of the white dwarf merger problem. It includes descriptions of 
the physics involved, the code options and runtime parameters, the diagnostic output 
and other I/O information, and the analysis routines. A full description of the physics 
and the associated numerical algorithms is included in the \castro\ code paper \cite{castro}.

@@

\castro\ includes full self-gravity with the Poisson equation, as well as the ability to perform a
simulation in a rotating reference frame with fixed rotation period. By default the $z$ axis will 
be the rotation axis, with the stars initially aligned along the $x$ axis. We select the 
orbital period of the system, and the masses of the \textit{primary} (more massive) and 
\textit{secondary} (less massive) white dwarfs. We then use Kepler's third law to determine the 
appropriate initial orbital distance. The center of mass of the system corresponds with the center 
of the domain, which is the origin; the domain is $1.024 \times 10^{10}\ \text{cm}$ wide in each 
coordinate direction, so that each dimension runs from $-5.12 \times 10^{9}\ \text{cm}$ to 
$+5.12 \times 10^{9}\ \text{cm}$. In a rotating reference frame, we set the stars at rest. 
We can instead run a simulation in the inertial reference frame. To make it equivalent, we
start all zones in the domain with a velocity corresponding to the equivalent rotating 
reference frame velocity. This means that the stars will be synchronously rotating (though we
may relax this assumption in future work).

Though we are mainly interested in 3D simulations, it is sometimes useful to compare to 2D
simulations which are much cheaper and can be run at higher resolutions. The only physically
meaningful analogue to these mergers in lower dimensions occurs in 2D axisymmetric cylindrical
coordinates. CASTRO represents these as $(r,z,\theta)$. In this coordinate system we enforce
that the stars lie along the $z$ axis and that the center of the stars lies at $r = 0$. The
$\theta$ dimension doesn't get directly evolved hydrodynamically but that is the dimension
along which the rotational motion occurs, so in some circumstances the velocity in that
direction will change with time because we still allow the $\theta$-velocity to participate in
the Coriolis force. When rotation is enabled, the rotation is along the $z = 0$ axis in the $r$
coordinate dimension. To try this out you should use \texttt{DIM = 2} in the makefile and
use \texttt{inputs\_2d} in the \texttt{wdmerger} directory in \castro.

@@

There are a few \castro\ input parameters that you will typically want to control when running problems. 
Among the most important is whether we are doing rotation (\texttt{castro.do\_rotation}) and if so, 
what the rotational period is (\texttt{castro.rotational\_period}). (Note that the rotational period 
is used to set the orbital properties even if we're not rotating, so you should always set this appropriately.)
You'll also want to change the \texttt{stop\_time}, which is when the simulation will terminate. 
You should also avoid flooding yourself with plotfiles and checkpoints by setting \texttt{amr.plot\_files\_output} and 
\texttt{amr.checkpoint\_files\_output}, respectively; these are the simulation time differences
between successive plotfiles and checkpoints. Note that \castro\ does not adjust the timestep 
to meet this criterion, so it will generate these at the first timestep that passes this interval.

@@

examining \texttt{probin}. (Some parameters in there are currently beta quality and 

@@

The \texttt{tagging} namelist is described in the AMR chapter of the \castro\ user's guide. At present, we 
only tag based on absolute values of the density and temperature.

@@

The routines that generate both of these are found in \\\texttt{sum\_integrated\_quantities.cpp}. 
This is a replacement of the \castro\ function that does something similar. We use the same 
conventions; for example, if summing over the whole grid and we encounter a region that is covered by 
a fine grid, we zero out the contribution from the coarse grid and only count the fine grid quantities 
in that location.

@@

At a minimum, your working directory needs a \texttt{compile/} subdirectory with a \texttt{makefile}
in it, and some \texttt{bash} script that you run to submit the jobs. If you want to build 
with the default settings, then your makefile needs only to have the line:
\begin{lstlisting}
  include $(WDMERGER_HOME)/source/GNUmakefile
\end{lstlisting}%$
This includes the main makefile from the \texttt{source/} directory in \texttt{wdmerger}. 
The main makefile sets all of our build options (compiler, number of space dimensions, 
equation of state, etc.) and then includes the \castro\ makefile, which then runs as normal.
If you want to adjust the build options, then simply add a line before this include. As an 
example, if you want to build using a gamma-law equation of state, set:
\begin{lstlisting}
  EOS_dir = gamma_law_general/

@@

Decide what directory you want to build the problem in. It is expected 
to be a subdirectory of the \texttt{results/} directory, which will be
automatically created for you:
\begin{lstlisting}
dir=results/test_problem/
\end{lstlisting}
Now, optionally decide how many processors you want to run with, and 
how long you want the job to run:
\begin{lstlisting}
nprocs=128
walltime=4:00:00
\end{lstlisting}
If you do not, the job submission script will default to running the job
using one node for one hour. Finally, run your job:
\begin{lstlisting}
run $dir $nprocs $walltime
\end{lstlisting}%$
That is all you need to submit a job. Return to the terminal, and type \texttt{bash run.sh}
and your job will execute. The helper function \texttt{run} does several things. First, 
it creates the directory named \texttt{\$dir}. Then, it copies into \texttt{\$dir} the \texttt{inputs}
and \texttt{probin} files from the root \texttt{wdmerger} source directory; these are the runtime
files \castro\ needs to determine what your problem setup will be. They contain the default 
parameters for merger runs. It also copies any auxiliary files one might need, like the 
table for the equation of state, as well as the actual \castro\ executable you just built in the 
\texttt{compile/} directory, and a copy of the run utilities that manage the job itself (so that the 
run doesn't change if you make local edits to the job management system). Finally, it builds a script 
that you'll use for actually submitting the job to the batch submission system on the machine you're on, 
and then submits that job. Thus for a typical setup, your run directory will contain at least the following 
files when you submit:
\begin{itemize}
  \item \texttt{Castro3d*.ex}
  \item \texttt{helm\_table.dat}
  \item \texttt{inputs}
  \item \texttt{probin}
  \item \texttt{run\_script}
  \item \texttt{job\_scripts/}
\end{itemize}
Note that if you want to set up this directory as above but \textit{not} actually submit the job, 
set the \texttt{no\_submit} variable to any value in your run script. Similarly, if you want to 
set up the directory to contain the inputs and probin file but none of the machine-dependent 
scripts or the executable, set the \texttt{inputs\_only} variable to any value in your run script.

@@

The \texttt{science/wdmerger\_I/} directory contains a number of example problems that we use
to assess the reliability of \castro\ in performing hydrodynamics simulations 
with self-gravity. This section includes a list of the existing problems 
as well as the instructions for building and running a problem.

@@

This project was started as the Ph.D. dissertation project for Max Katz at Stony Brook University. 
Michael Zingale is Max's primary thesis advisor. Correspondence and requests for information 
about this software should be directed to them; their e-mail addresses are maximilian.katz@stonybrook.edu
and michael.zingale@stonybrook.edu, respectively. Alan Calder and Doug Swesty at Stony Brook are 
also co-investigators on this project. The \castro\ code was developed primarily at the 
Lawrence Berkeley National Laboratory. Staff there that have contributed directly and 
indirectly to this project include Ann Almgren, John Bell, Weiqun Zhang, Andy Nonaka, and Vince Beckner.
